---
title: "Introduction to implementation steps of MetaCycle"
author: "Gang Wu, Ron Anafi, Michael Hughes, Karl Kornacker, and John Hogenesch"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to implementation steps of MetaCycle}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---
The **MetaCycle** package is mainly used for detecting rhythmic signals from large scale time-series data. Depending on features of each time-series data, **MetaCycle** incorporates [ARSER(ARS)](https://github.com/cauyrd/ARSER), [JTK_CYCLE(JTK)](http://openwetware.org/wiki/HughesLab:JTK_Cycle), and [Lomb-Scargle(LS)](http://research.stowers-institute.org/efg/2005/LombScargle/) properly for periodic signal detection, and it could also output integrated analysis results if required. 

This vignette introduces implementation of method selection and integration steps of **MetaCycle**, which are not detaily explained in help files. For looking at how to use two main functions--**meta2d** and **meta3d** of this package, please see 'Examples part' of each function's help file. 

The **MetaCycle** source code will be available on github later.

## Time-series datasets
### Two main categories
A typical time-series dataset from a non-human organism is a two-dimensional matrix. Each row indicates one molecular's profile along time, and all molecule at any one time point are detected from the same sample. It is usually not necessary to keep track of which individual organism does a sample come from. For easily explanation, we named this kind of dataset as 2D time-series dataset. Take the time-series transcriptome dataset from mouse liver as an example.

```{r}
library(MetaCycle)
head(cycMouseLiverRNA[,1:5])
```

For time-series datasets from human, it is usually essential to keep track of the individual information about each sample. Except one matrix stores experimental values of detected molecule from all samples, another matrix is necessary to store individual information of each sample. This kind of dataset is named as 3D time-series dataset. For example, a time-series dataset from human blood is shown as below. 

The individual information matrix:

```{r}
set.seed(100)
row_index <- sample(1:nrow(cycHumanBloodDesign), 4)
cycHumanBloodDesign[row_index,]
```

The corresponding experimental values:

```{r}
sample_id <- cycHumanBloodDesign[row_index,1]
head(cycHumanBloodData[,c("ID_REF", sample_id)])
```

A 3D time-series dataset could be divided into multiple 2D time-series datasets, and all experimental values for an individual under the same treatment are in one 2D time-series dataset. For example, we could extracted all experimental values from "AF0004" under "SleepExtension" into one 2D time-series dataset.

```{r}
group_index <- which(cycHumanBloodDesign[, "group"] == "SleepExtension")
cycHumanBloodDesignSE <- cycHumanBloodDesign[group_index,]
sample_index <- which(cycHumanBloodDesignSE[, "subject"] == "AF0004")
sample_AF0004 <- cycHumanBloodDesignSE[sample_index, "sample_library"]
cycHumanBloodDataSE_AF0004 <- cycHumanBloodData[, c("ID_REF", sample_AF0004)]
head(cycHumanBloodDataSE_AF0004)
```

### Detail types of 2D time-series dataset
One kind of usual 2D time-series dataset is evenly sampled once at each time point, and the interval between neighbour time points is integer. Not all datasets are as simple as this. There are datasets with replicate samples, or with missing values, or un-evenly sampled, or sampled with a non-integer interval. Examples of these types of dataset are shown in the below table. 

Data Type                    | Point 1| Point 2| Point 3| Point 4| Point 5| Point 6
---------------------------  | -------| -------| -------| -------| -------| -------
The usual data               | CT0    | CT4    | CT8    | CT12   | CT16   | CT20  
With missing value           | CT0    | NA     | CT8    | CT12   | CT16   | CT20  
With replicates              | CT0    | CT0    | CT8    | CT8    | CT16   | CT16  
With un-even interval        | CT0    | CT2    | CT8    | CT10   | CT16   | CT20
With non-integer interval    | CT0    | CT4.5  | CT9    | CT13.5 | CT18   | CT22.5  

Of course, some datasets may seem combination of two or more of above types of data.

Data Type                             | Point 1| Point 2| Point 3| Point 4| Point 5| Point 6
------------------------------------  | -------| -------| -------| -------| -------| -------
With replicates and missing value     | CT0    | CT0    | CT8    | NA     | CT16   | CT16  
With un-even interval and replicates  | CT0    | CT2    | CT2    | CT10   | CT16   | CT20

## Method selection
The **meta2d** function in **MetaCycle** is designed to analyze 2D time-series datasets, and it could automatically select proper method to analyze different types of input datasets. The implementation strategy used for **meta2d** is shown in the flow chart (drawn with "diagram" package).

```{r, echo=FALSE, warning=FALSE, fig.width=6.65, fig.height=5}
library(shape)
library(diagram)
par(mar = c(1, 0.5, 1, 0.5))
openplotmat()
#number of elements in each row
num_element <- c(1, 1, 2, 3, 4, 5)
#get position information of each element in the flow chart
elpos <- coordinates (num_element, mx = 0)
#adjust x-position of some elements
elposM <- elpos
elposM[1:2, 1] <- elposM[1:2, 1] - 0.21
elposM[4,1] <- elposM[4,1] - 0.269
elposM[7,1] <- elposM[7,1] - 0.186
elposM[11,1] <- elposM[11,1] - 0.076
#give information of strat and end point of each arrow
fromto <- c( c(1, 2), c(2,12), c(2,4), c(4,13), c(4, 7),
             c(7, 14), c(7,11), c(11, 15), c(11, 16) )
fromtoM <- matrix(ncol = 2, byrow = TRUE, data = fromto)
rownum <- nrow(fromtoM)
#draw arrow and store arrow position informaion in 'arrposM'
arrposM <- matrix(ncol = 2, nrow = rownum)
for (i in 1:rownum)
{
    arrposM[i, ] <- bentarrow (from = elposM[fromtoM[i, 1], ], to = elposM[fromtoM[i, 2], ], 
                               lcol = "blue", lwd = 1, arr.pos = 0.56, arr.length = 0.3, arr.lwd = 0.8)
}
#draw elements of flow chart
textparallel(mid = elposM[1,], radx = 0.089, rady=0.06, lab = "Time-series\ndata", lcol = "blue", 
             lwd=2, shadow.size = 0, cex = 0.72, font=2, theta=80)
textdiamond(mid =  elposM[2,], radx = 0.12, rady=0.066, lab = "With\nnon-integer\nintervals?", 
            lcol = "blue", lwd=2, shadow.size = 0, cex = 0.72, font=2)
diamond_index <- c(4, 7, 11)
diamond_lab <- c("Uneven\nsampling?", "Missing\nvalue?", "With\nreplicates?")
for (i in 1:length(diamond_index))
{
  textdiamond(mid = elposM[diamond_index[i],], radx = 0.08, rady=0.066, lab = diamond_lab[i], 
              lcol = "blue", lwd=2, shadow.size = 0, cex = 0.72, font=2)  
}

round_index <- 12:16
round_lab <- c("LS", "LS", "JTK&LS", "JTK&LS", "ARS&\nJTK&LS")
for (j in 1:length(round_index))
{
    textround(mid = elposM[round_index[j],], radx=0.056, rady=0.05, lab = round_lab[j],
              lcol = "blue", lwd=2, shadow.size = 0, cex =0.76, font=2, rx = 0.02)
}
#add 'Y' and 'N' on the flow chart
midposM <- elposM[c(2, 4, 7, 11),]
xpos <- midposM[,1]
ypos <- midposM[,2]
YposM <- cbind(c( (xpos[1] - 0.145), (xpos[2:3] - 0.1), (xpos[4] - 0.086) ), ypos + 0.02)
NposM <- cbind(c( (xpos[1] + 0.145), (xpos[2:3] + 0.1), (xpos[4] + 0.086) ), ypos + 0.02)
text(YposM[,1], YposM[,2], labels="Y", cex=0.8, font=2)
text(NposM[,1], NposM[,2], labels="N", cex=0.8, font=2)
```

For analyzing 3D time-series dataset, **meta3d** function in **MetaCycle** is suggested. It firstly divides the input dataset into multiple 2D time-series datasets based on individual information, and then use the defined method through calling **meta2d** to analyze each divided dataset. 

## Integration
In addition to selecting proper methods to analyze different kinds of datasets, **MetaCycle** could also output integrated results. In detail, **meta2d** integrates analysis results from multiple methods and **meta3d** integrates analysis results from multiple individuals.

### Pvalue
[Fisher's method](https://en.wikipedia.org/wiki/Fisher%27s_method) is implemented in both **meta2d** and **meta3d** for integrating multiple p-values. The below formula is used to combine multiple p-values into one test statistic (X^2^). 

$$X^2_{2k} \sim -2\sum_{i=1}^k ln(p_i)$$

X^2^ has a chi-squared distribution with 2k degrees of freedom (k is the number of p-values), when all the null hypotheses are true, and each p-value is independent. The combined p-value is determined by the p-value of X^2^.

### Period and phase
The integrated period from **MetaCycle** is an arithmetic mean value of multiple periods, while phase integration based on [mean of circular quantities](https://en.wikipedia.org/wiki/Mean_of_circular_quantities) is implemented in **meta2d** and **meta3d**. The detail steps are as below.

* convert phase values to polar coordinates $\alpha_j$
* convert polar coordinates to cartesian coordinates ($cos\alpha_j$, $sin\alpha_j$)
* compute the arithmetic mean of these points and its corresponding polar coordinate $\bar{\alpha}$
$$\bar{\alpha} = atan2(\frac{\sum_{j=1}^n sin\alpha_j}{n}, \frac{\sum_{j=1}^n cos\alpha_j}{n})$$
* convert the resulting polar coordinate to a integrated phase value

```{r, warning=FALSE}
# given three phases
pha <- c(0.9, 0.6, 23.6)
# their corresponding periods
per <- c(23.5, 24, 24.5)
# mean period length
per_mean <- mean(per)
# covert to polar coordinate
polar <- 2*pi*pha/per
# get averaged ploar coordinate
polar_mean <- atan2(mean(sin(polar)), mean(cos(polar)))
# get averaged phase value
pha_mean <- per_mean*polar_mean/(2*pi)
pha_mean
```

### Amplitude calculation
**meta2d** recalculates the amplitude with following model:

$$Y_i = B + TRE*(t_i - \frac{\sum_{i=1}^n t_i}{n}) + A*cos(2*\pi*\frac{t_i - PHA}{PER})$$

where $Y_i$ is the observed value at time $t_i$; B is baseline level of the time-series profile; TRE is trend level of the time-series profile; A is the amplitude of the waveform. PHA and PER are integrated period and phase mentioned above. In this model, only B, TRE and A are unknown parameters, which could be calculated with ordinary least square (OLS) method. The baseline and trend level are explained in the below example.

```{r, echo=FALSE, warning=FALSE, fig.width=6.65, fig.height=5}
getAMP <- function(expr, per, pha, tim=18:65)
{ 
    trendt <- tim - mean(tim[!is.na(tim) & !is.nan(tim)])
    cost <- cos(2*pi/per*(tim - pha))
    fit <- lm(expr~trendt + cost)
    fitcoef <- fit$coefficients
    basev <- fitcoef[1]
    trendv <- fitcoef[2]
    ampv <- fitcoef[3]
    fitexp <- basev + trendv*trendt + ampv*cost
    outL <- list("base"=basev, "trend"=trendv, "amp"=ampv, "fit"=fitexp)
    return(outL)
}

cirD <- cycVignettesAMP
ampL <- getAMP(expr=as.numeric(cirD[1,24:71]), per=cirD[1, "meta2d_period"], pha=cirD[1, "meta2d_phase"])

lay<-layout(cbind(1, 2), widths=c( lcm(cm(4.5)), lcm(cm(1.5)) ), heights=lcm(cm(4.5)) )
par(mai=c(0.65,0.6,0.4,0.05),mgp=c(2,0.5,0),tck=-0.01)
xrange <- c(18, 65)
yrange <- c(200, 2350)
plot(18:65, cirD[1,24:71], type="b", xlim=xrange, ylim=yrange, xlab="Circadian time(CT)", ylab="Expression value",  main=cirD[1,1], cex.main=1.2)
par(new=T)
plot(18:65, ampL[[4]], type="b", xlim=xrange, ylim=yrange, col="red", xlab="", ylab="", main="")
abline(h=ampL[[1]], lty=3, col="purple", lwd=1.5)
lines(18:65, 500+ampL[[2]]*(18:65-mean(18:65)), lty=4, col="orange", lwd=1.5)
legend("topleft", legend=c("Raw value", "OLS fitted value"), col=c("black", "red"), pch=1, bty="n")
legend("topright", legend=c("Baseline", "Trend"), col=c("purple", "orange"), lty=c(3, 4), lwd=1.5, bty="n" )

par(mai=c(0.5,0.05,0.4,0.1),mgp=c(2,0.3,0),tck=-0.01);
plot(x=NULL,y=NULL,xlim=c(0,10),ylim=c(0,10),type="n", xaxt="n",yaxt="n",bty="n",xlab="",ylab="",main="")
text(rep(1,3), c(8, 5, 2), c("Base = ", "Trend = ", "AMP = "), adj=0)
text_value <- unlist(ampL)
text(rep(6,6), c(8, 5, 2), round(text_value[1:3], 1), adj=0)
```

In addition, **meta2d** also output a relative amplitude value (rAMP), which could be easily taken as the ratio between amplitude and baseline (if |B| >= 1). The amplitude value is associated with the general expression level, which indicates highly expressed genes may always have larger amplitude than lowly expressed genes. The rAMP may be used to compare the amplitude values among genes with different expression levels. For example, *Ugt2b34* has a larger amplitude than *Arntl*, but its rAMP is smaller than *Arntl*.

```{r, echo=FALSE, warning=FALSE, fig.width=6.65, fig.height=5}
cirD <- cycVignettesAMP
cirM <- as.matrix(cirD[2:3, 24:71])
expmax <- apply(cirM, 1, max)
cirM <- cirM/expmax

lay<-layout(cbind(1, 2), widths=c( lcm(cm(4.5)), lcm(cm(1.5)) ), heights=lcm(cm(4.5)) )
par(mai=c(0.65,0.6,0.4,0.05),mgp=c(2,0.5,0),tck=-0.01)
xrange <- c(18, 65)
yrange <- c(0, 1)
colname <- c("red", "blue")
grey_trans <- rgb(191/255,191/255,191/255,0.65);

par(mai=c(0.65,0.6,0.2,0.05),mgp=c(2,0.5,0),tck=-0.01)
plot(NULL,NULL,xlim=xrange,ylim=yrange,xaxt="n",yaxt="n",xlab="Circadian time(CT)",ylab="Exp/Max", main="");
rect_xL <- c(18, 36, 60)
rect_yL <- c(24, 48, 65)
rect(rect_xL, rep(-0.1, 3), rect_yL, rep(1.1,3), col=grey_trans, border=NA, bty="n")

for (i in 1:2)
{
  loessD <- data.frame(expd=as.numeric(cirM[i,]),tp=18:65);
  exploess <- loess(expd~tp, loessD, span = 0.2);
  expsmooth <- predict(exploess, data.frame(tp=18:65));
  lines(18:65,expsmooth,lwd=1.2,col=colname[i]);
}

xpos <- c(seq(18,60,by=6), 65)
axis(side=1,at=xpos,labels=xpos,mgp=c(0,0.2,0),tck=-0.01,cex.axis=0.8)
ypos <- seq(0, 1, by=0.2)
axis(side=2,at=ypos,labels=ypos,mgp=c(0,0.2,0),tck=-0.01,cex.axis=0.8)

par(mai=c(0.5,0.05,0.2,0.1),mgp=c(2,0.3,0),tck=-0.01)
plot(x=NULL,y=NULL,xlim=c(0,10),ylim=c(0,10),type="n", xaxt="n",yaxt="n",bty="n",xlab="",ylab="",main="")
lines(c(0.2,2.3), c(9.5,9.5), col="blue", lwd=1.5)
text(c(5,0,0), c(9.5, 8, 6.5), c("Ugt2b34", "AMP = ", "rAMP = "), col="blue", adj=0)
text(c(5,5), c(8, 6.5), round(as.numeric(cirD[3, 22:23]), 2), col="blue", adj=0)

lines(c(0.2,2.3), c(4.5,4.5), col="red", lwd=1.5)
text(c(5,0,0), c(4.5, 3, 1.5), c("Arntl", "AMP = ", "rAMP = "), col="red", adj=0)
text(c(5,5), c(3,1.5), round(as.numeric(cirD[2, 22:23]), 2), col="red", adj=0)
```

Based on the calculated baseline, amplitude and relative amplitude values by **meta2d**, **meta3d** calculates the corresponding integrated values with arithmetic mean of multiple individuals in each group. 